Historical trends in the contribution of six major causes of death to the black-white

### 2017-05-16 Manuscript/Manuscript_COD_Trends.Rmd
manual bibliography

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref MMWR]. While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years [@harper2014trends]. However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat]. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data]. The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [@census_map]. We coulnd not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census. *Furthermore, the agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].*

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. First, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9 [ref NCHS or NCI page on suppression], we needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied [@preston; @selvin_2004]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method [@arriaga1984measuring]. This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher age-standardized rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate, say -0.5 years for injuries, this implies that whites have a higher age-standardized rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000 [ref census_stdpop].

@@

We examined state-level trends in six major cause of death contributions to the black-white life expectancy gaps over a 45 year period. While we found that many states conformed to the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. In contrast to some recent work focused on county-level inequalities [@arora2016population; @dwyer2016us], states have some substantial advantages as analysis units. They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.

Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century. [add: Lemann, 1991] Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.  On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy [@dwyer2016us]. On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends. [add: Case and Deaton, 2017]

Consistent with what has been reported at the national level[@harper2012trends], we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic [ref MMWR_opioids].

@@

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.  States pursue varying approaches to an array of relevant programs and laws including tobacco control [@cotti2016effects], Medicare coverage [@grogan2017racial], hand-gun violence [@humphreys2017evaluating] and nutrition [@taber2011state]. These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies]. The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.  Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[@strumpfMSE]

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group [@case2015rising; @minton2016two]. The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

Our analysis has two main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period [@laveist2011black].  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [@ruiz2013hispanic]. We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref]. In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [@harper2014trends], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall. As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient [@brazil2015spatial].

Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [@preston2014anatomy]. As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Such changes would both serve to increase the racial gap in DC [add: US Census bureau, 2015].

Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses [@kuehn2014cdc; @jones2013pharmaceutical].

@@

to add to bibliography list:
MMWR: Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH. Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015. MMWR Morb Mortal Wkly Rep. 2017 May 5;66(17):444-456. PMID: 28472021

seer stat: National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: latest release: version 8.1.5 [Internet]. Bethesda (MD): NCI; 2013 Mar 31 [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/seerstat

nchs data: National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/mortality/

cdc_hisp: [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf]

cdc_validity_race: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf

census_std_pop: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. retrieved apr 25, 2017

kochanek_2013: https://www.cdc.gov/nchs/data/databriefs/db125.pdf

Lemann, Nicholas. The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, 1991. ISBN 0-394-56004-3.

Case A, Deaton A. Mortality and morbidity in the 21st century. Brookings Working Paper, 2017.

Strumpf E, Harper S, Kaufman J. Fixed Effects and Difference-in-Differences. In: Methods in Social Epidemiology (Oakes & Kaufman, Eds) Jossey-Bass: John Wiley & Sons, 2017.

U.S. Census Bureau (2015). American Community Survey 1-year estimates. Retrieved from Census Reporter Profile page for Washington, DC <https://censusreporter.org/profiles/16000US1150000-washington-dc/>

MMWR opioids: https://www.cdc.gov/mmwr/volumes/65/wr/mm655051e1.htm Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015 Rose Rudd et al.

>>>>

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.^2^

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013.^5^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^6^ The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010.^7^ We could not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census. *Furthermore, the agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates.^8^*

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. First, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ we needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the gender-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^10, 11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher age-standardized rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate, say -0.5 years for injuries, this implies that whites have a higher age-standardized rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^

@@

We examined state-level trends in six major cause of death contributions to the black-white life expectancy gaps over a 45 year period. While we found that many states conformed to the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. In contrast to some recent work focused on county-level inequalities ,^14,15^ states have some substantial advantages as analysis units. They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.

Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century.^16^  Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.  On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^17^

Consistent with what has been reported at the national level,^18^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^19^

@@

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.  States pursue varying approaches to an array of relevant programs and laws including tobacco control,^20^ Medicare coverage,^21^ hand-gun violence,^22^  and nutrition.^23^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks.^24^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.  Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.^25^

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^26, 27^ The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

Our analysis has two main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period.^28^  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded.^29^ We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period.^8^ In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results,^4^ providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall. As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient.^30^

Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010.^31^ As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Such changes would both serve to increase the racial gap in DC.^32^

Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^33,34^

@@

1. Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH. Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015. MMWR Morb Mortal Wkly Rep. 2017;66(17):444-456. doi: 10.15585/mmwr.mm6617e1.

2. Fuchs, VR. Black Gains in Life Expectancy. JAMA. 2016;316(18): 1869–70. doi: 10.1001/jama.2016.14398.

3. Harper S, Lynch J, Burris S, Smith GD. Trends in the black-white life expectancy gap in the United States, 1983-2003. JAMA. 2007;297(11):1224-32.

4. Harper, Sam, Richard F MacLehose, and Jay S Kaufman. 2014. “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009.” Health Affairs 33 (8). Health Affairs: 1375–82.

5. National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.

6. Centers for Disease Control and Prevention. National Center for Health Statistics: National Vital Statistics System [Internet]. [cited 2017 May 16]. Available from: https://www.cdc.gov/nchs/nvss/index.htm.

7. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.

8. Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E. The validity of race and Hispanic origin reporting on death certificates in the United States. National Center for Health Statistics. Vital Health Stat 2(148). 2008. Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf

9. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2017 May 16]. Available from: http://www.seer.cancer.gov/mortality/

10. Preston SH, Heuveline P, and Guillot M. Demography: Measuring and Modeling Population Processes. Malden: Blackwell Publishing; 2001.

11. Selvin, S. Statistical Analysis of Epidemiologic Data. 3rd ed. Oxford University Press; 2004.

12. Arriaga EE. Measuring and explaining the change in life expectancies. Demography. 1984;21(1):83-96.

13. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet]. [cited 2017 April 25]. Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.

14. Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM. Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level. Health Affairs. 2016;35(11):2075-82.

15. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al. US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. JAMA. 2016;316(22):2385–2401.

16. Lemann, N. The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, 1991. ISBN 0-394-56004-3.

17. Case A, Deaton A. Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017; 2017.

18. Harper S, Rushani D, and Kaufman JS. Trends in the Black-White Life Expectancy Gap, 2003-2008. JAMA. 2012;307(21):2257–9.

19. Rudd RA, Seth P, David F, Scholl L. Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. MMWR Morb Mortal Wkly Rep 2016;65:1445–1452. DOI: http://dx.doi.org/10.15585/mmwr.mm655051e1.

20. Cotti C, Nesson E, Tefft N. The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data. American Economic Journal: Economic Policy. 2016;8(4):103-23.

21. Grogan CM, Park SE. The Racial Divide in State Medicaid Expansions. Journal of Health Politics, Policy and Law. 2017 Jun 1;42(3):539-72.

22. Humphreys DK, Gasparrini A, Wiebe DJ. Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study. JAMA internal medicine. 2017;177(1):44-50.

23. Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC. State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption. American Journal of Public Health. 2011;101(9):1769-75.

24. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS. Policies and politics that promote HIV infection in the Southern United States. AIDS (London, England). 2014;28(10):1393.

25. Strumpf EC, Harper S, Kaufman JS. Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed. (Oakes & Kaufman, Eds) San Francisco: Jossey-Bass, 2017.

26. Case A, Deaton A. Rising Morbidity and Mortality in Midlife among White Non-Hispanic Americans in the 21st Century. Proceedings of the National Academy of Sciences. 2015 Dec 8;112(49):15078-83.

27. Minton J, Green M, McCartney G, Shaw R, Vanderbloemen L, Pickett K. Two cheers for a small giant? Why we need better ways of seeing data: A commentary on:‘Rising morbidity and mortality in midlife among White non-Hispanic Americans in the 21st century’. International Journal of Epidemiology. 2016 Sep 17:dyw095.

28. LaVeist-Ramos TA, Galarraga J, Thorpe RJ, Bell CN, Austin CJ. Are black Hispanics black or Hispanic? Exploring disparities at the intersection of race and ethnicity. Journal of Epidemiology and Community Health. 2011 Mar 3:jech-2009.

29. Ruiz JM, Steffen P, Smith TB. Hispanic mortality paradox: a systematic review and meta-analysis of the longitudinal literature. American Journal of Public Health. 2013 Mar;103(3):e52-60.

30. Brazil N. Spatial Variation in the Hispanic Paradox: Mortality Rates in New and Established Hispanic US Destinations. Population, Space and Place. 2017;23:e1968. DOI:10.1002/psp.1968.

31. Preston SH, Elo IT. Anatomy of a municipal triumph: New York City's upsurge in life expectancy. Population and development review. 2014;40(1):1-29.

32. U.S. Census Bureau. Census Reporter Profile Page: American Community Survey 1-year estimates [Internet]. [Cited 16 May 2017]. Available from: https://censusreporter.org/profiles/16000US1150000-washington-dc/.

33. Kuehn BM. CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. JAMA. 2014;312(7):684-6.

34. Jones CM, Mack KA, Paulozzi LJ. Pharmaceutical overdose deaths, United States, 2010. JAMA. 2013;309(7):657-9.
